---
title: "Synthetic Control Analysis"
author: "Asia Kaiser"
date: "2024-10-17"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(tidyverse)
library(tidysynth)
library(nlme)
library(AICcmodavg)
library(purrr)
library(ggpubr)
library(microsynth)
```

# Load in Data
```{r}
rm(list = ls())
city.data <- read.csv("data/cities.scm.input.csv")
```

DELETE THIS CHUNK AFTER RERUNNING 'GBIF DATA EXTRACTION SCRIPT'
```{r}
#Subset for cities with closer number of observations during treatment year
cities.pool <- city.data %>%
  filter(year == 2021) %>%
  filter(n.obs45km > 75 & n.obs45km < 500)

#Extract the city column
cities.pool <-cities.pool$City

#Keeping subset of cities in final dataframe
#Removing Boston, Atlanta and Baltimore for experiencing this treatment. Removing Houston for abnormal abundance spike in 2016.
city.data <- city.data %>%
  filter(City != "Boston"& City != "Atlanta" & City != "Baltimore" & City != "Houston")%>%
  filter(City %in% cities.pool)
```

# Synthetic Control Analysis: Abundance

## Creating the Synthetic Control
* Creating the control objects
* Selecting the predictors
* Generating weights

```{r}
# Create synthetic control object
gbifcity_out <- city.data %>%
  # Initial synthetic control object
  synthetic_control(outcome = n.obs45km, # outcome
                    unit = City, # unit index in the panel data
                    time = year, # time index in the panel data
                    i_unit = "Philadelphia", # unit where the intervention occurred
                    i_time = 2021, # time period when the intervention occurred
                    generate_placebos = TRUE # generate placebo synthetic controls (for inference)
  ) %>%
  
  #Generate aggregate predictors to fit weights to each unit
  generate_predictor(time_window = 2015:2023,
                     Temperature = mean(Temp,rm.na= TRUE),
                     Precipitation = mean(Prec, rm.na= TRUE),
                     Population = mean(Population, rm.na= TRUE),
                     ) %>%
  
  generate_predictor(time_window = 2021,
                     "Tree canopy cover" = tcc45km,                     ,
                     "Impervious surface" = imperv45km,
                     "Area (km2)" = area_km2
                     ) %>%

  # Generate the fitted weights for the synthetic control
  generate_weights(optimization_window = 2015:2021, # time to use in the optimization task
                   margin_ipop = .02,sigf_ipop = 7,bound_ipop = 6 # optimizer options
  ) %>%
  
  # Generate the synthetic control
  generate_control()
```


Creating Synthetic Control with MicroSynth to Extract Confidence Intervals
```{r}
#Creating a variable for treatment
city.data <- city.data |>
    mutate(
        Treatment = ifelse(City == "Philadelphia",1,0)
    )
```


*adjust the following code for my data*
#```{r}
sea1 <- microsynth(seattledmi, 
                   idvar="ID", timevar="time", intvar="Intervention", 
                   start.pre=1, end.pre=12, end.post=16, 
                   match.out=match.out, match.covar=cov.var, 
                   result.var=match.out, omnibus.var=match.out,
                   test="lower",
                   n.cores = min(parallel::detectCores(), 2))
sea1
#```


## Visualizing the data

Creating Personal Theme
```{r}
Asia_Theme <- theme(
  plot.title = element_text(size = 20),
  plot.caption = element_text(size = 16),
  axis.title = element_text(size = 20),
  axis.text = element_text(size = 20),
  axis.text.x = element_text(size = 20),
  legend.text = element_text(size= 20))
```


**Trends**: plot of trends of observed (treated) unit compared to synthetic control over time, with dashed vertical line representing the treatment event. 
```{r}
trendplot <- gbifcity_out %>% plot_trends()+
  Asia_Theme +
  labs(title = "Synthetic Control Method", caption = NULL,
        x = "Year",
        y = "Bee observations")
trendplot
```

**Differences**: plot of difference in outcome variable between the observed (treated) unit and the estimated outcome of the synthetic control.
```{r}
diffplot <- gbifcity_out %>% plot_differences()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
        x = "Year",
        y = "(Estimated - Observed) Bee observations")
```

**Weights**: Barplots showing the relative contributions (weights) of each control unit and each predictor/variable unit to the synthetic control.
```{r}
weightplot <- gbifcity_out %>% plot_weights()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
       y = "Weight")
weightplot
```

**Balance Table**: A table with values of the predictors/variables so that you can see how similar the synthetic control is to observed variables of the treated unit.
```{r}
balancetab <- gbifcity_out %>% grab_balance_table()
```

## Inference

**In-space placebos**: The same as difference plot above, this time comparing each control unit to the synthetic control. This is to see if the difference between the treated unit and the synthetic control actually stands out as significantly greater. 
```{r}
placebos <- gbifcity_out %>% plot_placebos()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
        x = "Year",
        y = "(Estimated - Observed) Bee observations")
placebos
```

**Ratio of pre and post intervention period mean squared predictive error (mspe)**
```{r}
mpseratio <- gbifcity_out %>% plot_mspe_ratio()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL)
mpseratio
```

### Significance table
**Constructing Fisher's Exact P-value**: This table takes the mspe ratio and lists units in descending order. P-values are constructed by dividing a units rank by the total. *You need at least 20 cases in the unit pool to achieve a p-value below .05*

```{r}
significance.tab <- gbifcity_out %>% grab_significance()
significance.tab <- as.data.frame(significance.tab)

#This code provides the causal estimate value (difference in outcome between the synthetic control and the observed treated unit)
gbifcity_out %>% grab_synthetic_control()%>%
  filter(time_unit == 2022)%>%
  mutate(causal_estimate= synth_y - real_y)%>%
  mutate(percent_estimate= (synth_y - real_y)/synth_y)%>%
  head
```

# Other Approaches

## Difference in Difference w/ most similar city
```{r}
#Creating the relevant dummy variables
city.data.did <- city.data %>%
  mutate(Time = ifelse(year >= 2021, 1, 0)) %>%
  mutate(Treated = if_else(City == "Philadelphia",1,0))%>%
    filter(City == "Philadelphia" | City == "Chicago")

#Running the Difference in Difference Regression
did.model <- lm(n.obs45km ~ Treated*Time, data = city.data.did)
summary(did.model)
confint(did.model)


prediction_intervals <- predict(
  did.model, 
  newdata = city.data.did, 
  interval = "confidence", 
  level = 0.95
  )

city.data.did <- cbind(city.data.did, prediction_intervals)
```

*Difference in Difference plot with most similar city*
```{r}
palette.colors(palette = "Okabe-Ito")

did.trendplot <- ggplot(city.data.did, aes(x = year, y = n.obs45km, color = City)) +
    geom_line(linewidth = 1, aes(linetype = City))+
    geom_ribbon(aes(ymin = lwr, ymax = upr, fill = City), alpha = 0.2, linetype = 0)+
  geom_point()+
  scale_linetype_manual(values=c(1,3))+
  scale_color_manual(values=c('#D55E00','#56B4E9'))+
  geom_vline(xintercept = 2021, linetype="dashed")+
  labs(title = "Difference in Difference",
       x = "Year",
       y = "Bee observations") +
  theme_minimal()+
  theme(legend.position="bottom",legend.title = element_blank())+
  Asia_Theme

did.trendplot
```

## Interrupted Time Series (ITS) Regression

Explanation of ITS modelling: 
https://rpubs.com/chrissyhroberts/1006858 
```{r}
#Creating the relevant dummy variables in dataframe
city.data.its <- city.data %>%
  filter(City== "Philadelphia" & year >= 2017)%>%
  mutate(Treated = if_else(year >= 2021, 1, 0))%>%
  mutate(Time_since= if_else(year < 2021, 0, (year-2020)))
city.data.its

#Running the its regression
its.model <- gls(n.obs45km ~ year + Treated + Time_since, data = city.data.its, correlation= corARMA(p=1, q=1, form = ~ year), method = "ML")
summary(its.model)

#confidence intervals for the its regression
confint(its.model)
```


ITS Plot
```{r}
#Creating prediction intervals
city.data.its <-city.data.its %>% mutate(
  its.predictions = predictSE.gls (its.model,city.data.its, se.fit=T)$fit,
  its.se = predictSE.gls (its.model,city.data.its, se.fit=T)$se
  )

#Creating model for counterfactual
city.data.its.2 <-filter(city.data.its, year < 2021)
its.counter = gls(n.obs45km ~ year, data = city.data.its.2, correlation= corARMA(p=1, q=1, form = ~ year),method="ML")

#Adding prediction intervals for counterfactual back to dataframe
city.data.its <-city.data.its %>% mutate(
  counter.predictions = predictSE.gls (its.counter, newdata = city.data.its, se.fit=T)$fit,
  counter.se = predictSE.gls (its.counter, city.data.its, se.fit=T)$se
)
```

Plot
```{r}
its.plot <- ggplot(city.data.its,
         aes(year,n.obs45km))+
  geom_ribbon(aes(ymin = counter.predictions - (1.96*counter.se), ymax = counter.predictions + (1.96*counter.se)), fill = '#D55E00', alpha = 0.2, linetype = 2)+
  geom_line(aes(year,counter.predictions),color='#D55E00', lty = 3)+
  geom_ribbon(aes(ymin = its.predictions - (1.96*its.se), ymax = its.predictions + (1.96*its.se)), fill = '#56B4E9', alpha = 0.2, linetype = 1)+
  geom_line(color='#56B4E9', lty = 1)+
    geom_vline(xintercept = 2021, linetype="dashed")+
  geom_point()+
      labs(title = "Interrupted time series",
       x = "Year",
       y = "Bee observations")+
    theme_minimal()+
    Asia_Theme
its.plot
```
#Confidence Intervals

Confidence interval plot for effect sizes in 2022 for all approaches
```{r}
#Put coefficient and confidence interval estimates in single dataframe
method <- c("Synthetic Control","Difference in Difference","Interrupted Time Series")
estimate <- c(40.3,did.model$coefficients[4],its.model$coefficients[3])
lwrCI <- c(70,confint(did.model)[4],confint(its.model)[3]) #replace SCM CI
uprCI <- c(10,confint(did.model)[8],confint(its.model)[7]) #replace SCM CI
ci.data <- tibble(methods,estimate,lwrCI,uprCI)
```
CI Forest plot
```{r}
ci.plot <- ggplot(ci.data,
       aes(x = estimate, y = method))+
    geom_errorbar(
        aes(xmin = lwrCI, xmax = uprCI), width = 0.25)+
    geom_point(size = 2.5)+
    geom_vline(xintercept = 0, linetype = 2)+
    theme_minimal()+
    Asia_Theme
ci.plot
    
```


# Save Figures
```{r}
trends.figure <- ggarrange(trendplot,did.trendplot,
                    labels = c("A", "B"))
ggsave("figures/ConfidenceIntervals_plot.png", plot = ci.plot)
ggsave("figures/Trend plot.png", plot = trendplot)
ggsave("figures/Difference plot.png", plot = diffplot)
ggsave("figures/Weight plot.png", plot = weightplot)
ggsave("figures/Placebos plot.png", plot = placebos)
ggsave("figures/MPSE ratio plot.png", plot = mpseratio)
ggsave("figures/Abundance estimates trend plot.png", plot = did.trendplot)
ggsave("figures/ITS_plot.png", plot = its.plot)
ggsave("figures/Method comparison plot.png", plot = trends.figure, height = 6, width = 12)
```

# Citations
```{r}
citation()
devtools::session_info()

c("tidyverse", "tidysynth") %>%
  map(citation) %>%
  print(style = "text")
```



