---
title: "Synthetic Control Analysis"
author: "Asia Kaiser"
date: "2024-10-17"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(tidyverse)
library(tidysynth)
library(nlme)
library(AICcmodavg)
library(purrr)
library(ggpubr)
library(microsynth)
library(fixest)
library(lmerTest)
```

# Load in Data
```{r}
rm(list = ls())
city.data <- read.csv("data/cities.scm.input.csv")
```

# Synthetic Control Analysis in scpi Package
```{r}
#write.csv(city.data, "city_dat.csv", row.names = FALSE)
city.data <- city.data %>%
    rename("nobs45km" = "n.obs45km")

# try in scpi
library(scpi)
library(data.table)

donor.cities <- unique(city.data$City)
donor.cities <- donor.cities[!(donor.cities == "Philadelphia")]
city.data$Temp <-scale(city.data$Temp)
city.data$Prec <- scale(city.data$Prec)
city.data$imperv45km <- scale(city.data$imperv45km)
city.data$Population <- scale(city.data$Population)
city.data$tcc45km <- scale(city.data$tcc45km)
city.data$area_km2 <- scale(city.data$area_km2)

feature.vars <- c("Temp","Prec","imperv45km","Population","nobs45km","tcc45km","area_km2")
city.data <- city.data %>%
  rename(unit.no = ID)


dat <- scdata(df = city.data,
              id.var = "City",
              time.var = "year",
              outcome.var = "nobs45km",
              period.pre = (2015:2021),
              period.post = (2022:2023),
              unit.tr = "Philadelphia",
              unit.co = donor.cities,
              features = feature.vars,
              cov.adj = list('Temp' = c(),
                             'Prec' = c(),
                             'imperv45km' = c(),
                             'Population' = c("constant","trend"),
                             'nobs45km' = c("constant","trend"),
                             'tcc45km' = c(),
                             'area_km2' = c()))
summary(dat)

est.si  <- scest(data = dat)
summary(est.si)

#Estimating uncertainty
sims <- 1000
u.order <- 1
u.lags <- 0
u.sigma <- "HC1"
u.missp <- TRUE
e.order <- 1
e.lags <- 0
e.method <- "gaussian"
lgapp <- "linear"
cores <- 1
set.seed(0)
res.pi <- scpi(data = dat, sims = sims, e.method = e.method, e.order = e.order, e.lags = e.lags, u.order = u.order, u.lags = u.lags, u.sigma = u.sigma, u.missp = u.missp, cores = cores, w.constr = list(name = "simplex"))
?scpi
summary(res.pi)

scplot(res.pi, e.out = TRUE)

nobs45km <- as.numeric(c(res.pi$data$Y.pre[,1],
                                   res.pi$data$Y.post[,1],
                                   res.pi$est.results$Y.pre.fit[,1],
                                   res.pi$est.results$Y.post.fit[,1]))
year <- rep(2015:2023,2)
Treatment <- c(rep("Philadelphia",9), rep("Synthetic Control",9))

scpi.data <- tibble(nobs45km,year,Treatment)

(decline.2023 <- ((329.979- 299)/329.979)*100)
(decline.2022 <- ((283.317-238)/283.317)*100) 
```

GGplot of Scpi estimates

Creating Personal Theme
```{r}
Asia_Theme <- theme(
  plot.title = element_text(size = 20),
  plot.caption = element_text(size = 16),
  axis.title = element_text(size = 20),
  axis.text = element_text(size = 20),
  axis.text.x = element_text(size = 20),
  legend.text = element_text(size= 20))
```

```{r}
scpi.plot <- ggplot(scpi.data, aes(x = year, y = nobs45km, color = Treatment)) +
    geom_line(linewidth = 1, aes(linetype = Treatment))+
  geom_point()+
    geom_point(aes(x = 2022, y = 283.317, color = "#c443b2", group = 1))+
        geom_errorbar(aes(x = 2022, ymin = 268.804, ymax = 328.980, color = "#c443b2"), width = 0.25)+
    geom_point(aes(x = 2023, y = 329.979, color = "#c443b2", group = 1))+
    geom_errorbar(aes(x = 2023, ymin = 321.669, ymax = 376.095, color = "#c443b2"), width = 0.25)+
  scale_linetype_manual(values=c(1,4,1))+
  scale_color_manual(values=c("#c443b2","darkgrey","#c443b2"))+
  geom_vline(xintercept = 2021, linetype="dashed")+
  labs(title = "Synthetic Control (SC)",
       x = "Year",
       y = "Bee observations") +
  theme_minimal()+
  theme(legend.position = "none")+
  Asia_Theme+
    ylim(0,425)+
    xlim(2015,2023.25)
scpi.plot
```


# Synthetic Control Analysis in tidysynth Package

## Creating the Synthetic Control
* Creating the control objects
* Selecting the predictors
* Generating weights

# Reload in the Data
```{r}
city.data <- read.csv("data/cities.scm.input.csv")
```


Tidysynth 
```{r}
# Create synthetic control object
gbifcity_out <- city.data %>%
  # Initial synthetic control object
  synthetic_control(outcome = n.obs45km, # outcome
                    unit = City, # unit index in the panel data
                    time = year, # time index in the panel data
                    i_unit = "Philadelphia", # unit where the intervention occurred
                    i_time = 2021, # time period when the intervention occurred
                    generate_placebos = TRUE # generate placebo synthetic controls (for inference)
  ) %>%
  
  #Generate aggregate predictors to fit weights to each unit
  generate_predictor(time_window = 2015:2023,
                     Population = mean(Population, rm.na= TRUE),
                     ) %>%

  generate_predictor(time_window = 2021,
                     "Temperature" = Temp,
                     "Precipitation" = Prec,
                     "Tree canopy cover" = tcc45km,                     ,
                     "Impervious surface" = imperv45km,
                     "Area (km2)" = area_km2
                     ) %>%

  # Generate the fitted weights for the synthetic control
  generate_weights(optimization_window = 2015:2021, # time to use in the optimization task
                   margin_ipop = .02,sigf_ipop = 7,bound_ipop = 6 # optimizer options
  ) %>%
  
  # Generate the synthetic control
  generate_control()
```


## Visualizing the data

**Trends**: plot of trends of observed (treated) unit compared to synthetic control over time, with dashed vertical line representing the treatment event. 
```{r}
trendplot <- gbifcity_out %>% plot_trends()+
  Asia_Theme +
  labs(title = "Synthetic Control (SC)", caption = NULL,
        x = "Year",
        y = "Bee observations")+
    ylim(0,425)+
    xlim(2015,2023)+
    theme(legend.position = "none")
trendplot
```

**Differences**: plot of difference in outcome variable between the observed (treated) unit and the estimated outcome of the synthetic control.
```{r}
diffplot <- gbifcity_out %>% plot_differences()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
        x = "Year",
        y = "(Estimated - Observed) Bee observations")
```

**Weights**: Barplots showing the relative contributions (weights) of each control unit and each predictor/variable unit to the synthetic control.
```{r}
weightplot <- gbifcity_out %>% plot_weights()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
       y = "Weight")
weightplot
```

**Balance Table**: A table with values of the predictors/variables so that you can see how similar the synthetic control is to observed variables of the treated unit.
```{r}
balancetab <- gbifcity_out %>% grab_balance_table() %>%
    mutate(synthetic_percent_diff = ((Philadelphia - synthetic_Philadelphia)/Philadelphia)*100,
           donor_percent_diff = ((Philadelphia - donor_sample)/Philadelphia)*100)
```

## Inference

**In-space placebos**: The same as difference plot above, this time comparing each control unit to the synthetic control. This is to see if the difference between the treated unit and the synthetic control actually stands out as significantly greater. 
```{r}
placebos <- gbifcity_out %>% plot_placebos(prune = FALSE)+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL,
        x = "Year",
        y = "(Estimated - Observed) Bee observations")
placebos
```

**Ratio of pre and post intervention period mean squared predictive error (mspe)**
```{r}
mpseratio <- gbifcity_out %>% plot_mspe_ratio()+
  Asia_Theme +
  labs(title = NULL, subtitle = NULL)
mpseratio
```

### Significance table
**Constructing Fisher's Exact P-value**: This table takes the mspe ratio and lists units in descending order. P-values are constructed by dividing a units rank by the total. *You need at least 20 cases in the unit pool to achieve a p-value below .05*

```{r}
significance.tab <- gbifcity_out %>% grab_significance()
significance.tab <- as.data.frame(significance.tab)

#This code provides the causal estimate value (difference in outcome between the synthetic control and the observed treated unit)
gbifcity_out %>% grab_synthetic_control()%>%
  filter(time_unit == 2022)%>%
  mutate(causal_estimate= synth_y - real_y)%>%
  mutate(percent_estimate= (causal_estimate/synth_y)*100)%>%
  head
```

# Other Approaches

## Before-After (BA) Analysis w/ closest city with similar population size (naive selection)

```{r}
#Create relevant dummy variables
city.data.ba <- city.data %>%
    filter(City == "Philadelphia")%>%
    mutate(Treatment = ifelse(year > 2021,1,0))

#ba.model <- feols(nobs45km ~ Treatment, data = city.data.ba, cluster = ~City)

ba.model <- lm(nobs45km ~ Treatment, data = city.data.ba)
summary(ba.model)

confint.ba <- confint(ba.model)
confint.ba
```


*BA plot with most similar city*
```{r, warning = FALSE}
ba.plot <- ggplot(city.data.ba, aes(x = year, y = nobs45km)) +
    geom_line(linewidth = 1, color = "darkgrey")+
  geom_point(color = "darkgrey")+
    geom_point(aes(x = 2018, y = ba.model$coefficients[1], color = "#8B0000", group = 1))+
        geom_errorbar(aes(x = 2018, ymin = 0, ymax = confint.ba[3], color = "#8B0000"), width = 0.25)+
    geom_point(aes(x = 2022, y = ( ba.model$coefficients[1] + ba.model$coefficients[2]), color = "#8B0000", group = 1))+
    geom_errorbar(aes(x = 2022, ymin = ( ba.model$coefficients[1] + confint.ba[2]), ymax = ( ba.model$coefficients[1] + confint.ba[4]), color = "#8B0000"), width = 0.25)+
    geom_segment(aes(x = 2018, y = ba.model$coefficients[1], xend = 2022, yend =  ( ba.model$coefficients[1] + ba.model$coefficients[2]), colour = "#8B0000"))+
  geom_vline(xintercept = 2021, linetype="dashed")+
  labs(title = "Before-After (BA)",
       x = "Year",
       y = "Bee observations") +
  theme_minimal()+
  theme(legend.position = "none")+
  Asia_Theme+
    ylim(0,425)+
    xlim(2015,2023)
ba.plot
```


##Before- After Control Impact (BACI) Analysis w/ closest city with similar population size (naive selection)
```{r}
#Creating the relevant dummy variables
city.data.baci <- city.data %>%
  mutate(Time = ifelse(year > 2021, 1, 0)) %>%
  mutate(Treated = if_else(City == "Philadelphia",1,0))%>%
    filter(City == "Philadelphia" | City == "Detroit")

#Running the Before- After Control Impact model
#baci.model <- feols(nobs45km ~ Treated * Time, data = city.data.baci, cluster = ~City)
baci.model <- lm(nobs45km ~ Treated * Time, data = city.data.baci)

summary(baci.model)
confint.baci <- confint(baci.model)
confint.baci
#shows significant increases in Philadelphia bee abundance when compared to Detroit
```

*BACI plot with most similar city*
```{r, warning=FALSE}
palette.colors(palette = "Okabe-Ito")

baci.plot <- ggplot(city.data.baci, aes(x = year, y = nobs45km, color = City)) +
    geom_line(linewidth = 1, aes(linetype = City))+
  geom_point()+
  scale_linetype_manual(values=c(3,1))+
  scale_color_manual(values=c('#D55E00','darkgrey'))+
  geom_vline(xintercept = 2021, linetype="dashed")+
  labs(title = "Before-After Control Impact (BACI)",
       x = "Year",
       y = "Bee observations") +
  theme_minimal()+
  theme(legend.position = "none")+
  Asia_Theme+
    ylim(0,425)+
    xlim(2015,2023)

baci.plot
```

Before- After Control Impact (BACI) Analysis w/ Cleveland
```{r}
#Creating the relevant dummy variables
city.data.baci.2 <- city.data %>%
  mutate(Time = ifelse(year > 2021, 1, 0)) %>%
  mutate(Treated = if_else(City == "Philadelphia",1,0))%>%
    filter(City == "Philadelphia" | City == "Cleveland")

#Running the Before- After Control Impact model
#baci.model <- feols(nobs45km ~ Treated * Time, data = city.data.baci, cluster = ~City)
baci.model.2 <- lm(nobs45km ~ Treated * Time, data = city.data.baci.2)

summary(baci.model.2)
confint.baci.2 <- confint(baci.model.2)
confint.baci.2
#shows significant increases in Philadelphia bee abundance when compared to Detroit
```

*BACI plot with Cleveland*
```{r, warning=FALSE}
palette.colors(palette = "Okabe-Ito")

baci.plot.2 <- ggplot(city.data.baci.2, aes(x = year, y = nobs45km, color = City)) +
    geom_line(linewidth = 1, aes(linetype = City))+
  geom_point()+
  scale_linetype_manual(values=c(3,1))+
  scale_color_manual(values=c('#D55E00','darkgrey'))+
  geom_vline(xintercept = 2021, linetype="dashed")+
  labs(title = "Before-After Control Impact (BACI)",
       x = "Year",
       y = "Bee observations") +
  theme_minimal()+
  Asia_Theme+
    ylim(0,425)+
    xlim(2015,2023)

baci.plot.2
```

## Interrupted Time Series (ITS) Regression

Explanation of ITS modelling: 
https://rpubs.com/chrissyhroberts/1006858 

```{r}
#Creating the relevant dummy variables in dataframe
city.data.its <- city.data %>%
  filter(City == "Philadelphia" & year >= 2015)%>%
  mutate(Treated = if_else(year > 2021, 1, 0))%>%
  mutate(Time_since= if_else(year <= 2021, 0, (year-2020)))
city.data.its

#Running the its regression
its.model <- gls(nobs45km ~ year + Treated + Time_since, data = city.data.its, correlation= corARMA(p=1, q=1, form = ~ year), method = "ML")
summary(its.model)

#confidence intervals for the its regression
confint.its <- confint(its.model)
```


ITS Plot - making predictions
```{r}
#Creating model for counterfactual
city.data.its.2 <-filter(city.data.its, year <= 2021)
its.counter = gls(nobs45km ~ year, data = city.data.its.2, correlation= corARMA(p=1, q=1, form = ~ year),method="ML")

#Adding prediction intervals for counterfactual back to dataframe
city.data.its <-city.data.its %>% mutate(
  counter.predictions = predictSE.gls (its.counter, newdata = city.data.its, se.fit=T)$fit,
  counter.se = predictSE.gls (its.counter, city.data.its, se.fit=T)$se
)
```

Plot
```{r}
its.plot <- ggplot(city.data.its,
         aes(x = year, y = nobs45km))+
  geom_ribbon(aes(ymin = counter.predictions - (1.96*counter.se), ymax = counter.predictions + (1.96*counter.se)), fill = 'turquoise', alpha = .2, linetype = 2)+
  geom_line(aes(year,counter.predictions),color='turquoise', lty = 4)+
  geom_point(color = 'darkgrey')+
    geom_line(color = 'darkgrey', linewidth = 1)+
      scale_linetype_manual(values=1)+
    geom_vline(xintercept = 2021, linetype="dashed")+
    xlim(2015,2023)+
    ylim(0,425)+
      labs(title = "Interrupted Time Series (ITS)",
       x = "Year",
       y = "Bee observations")+
    theme_minimal()+
    Asia_Theme
its.plot
```
#Confidence Intervals

Confidence interval plot for effect sizes in 2022 for all approaches
```{r}
#Put coefficient and confidence interval estimates in single dataframe
method <- c("BA","BACI","ITS")
estimate <- c(ba.model$coefficients[2],baci.model$coefficients[4],its.model$coefficients[3])
lwrCI <- c(confint.ba[2],confint.baci[4],confint.its[3]) #replace SCM CI
uprCI <- c(confint.ba[4],confint.baci[8],confint.its[7])
ci.data <- tibble(method,estimate,lwrCI,uprCI)
```

CI Forest plot - Alternative methods
```{r}
ci.plot <- ggplot(ci.data,
       aes(x = estimate, y = method))+
    geom_errorbar(
        aes(xmin = lwrCI, xmax = uprCI), width = 0.25)+
    geom_point(size = 2.5)+
    geom_vline(xintercept = 0, linetype = 2)+
    xlim(-160,350)+
    theme_minimal()+
    Asia_Theme+
    theme(axis.title.y = element_blank())
ci.plot
```

CI plot - synthetic control method
```{r}
method <- "   SC"
estimate <- (299 - 329.979) 
lwrCI <- (299 - 376.095)
uprCI <- (299 - 321.669)
scm.ci.data <- tibble(method,estimate,lwrCI,uprCI)

scm.ci.plot <- ggplot(scm.ci.data,
       aes(x = estimate, y = method))+
    geom_errorbar(
        aes(xmin = lwrCI, xmax = uprCI), width = 0.25)+
    geom_point(size = 2.5)+
    geom_vline(xintercept = 0, linetype = 2)+
    xlim(-160,350)+
    theme_minimal()+
    Asia_Theme+
    theme(axis.title.y = element_blank())
scm.ci.plot
```


# Save Figures
```{r,warning=FALSE}
trends.figure <- ggarrange(its.plot,baci.plot,ba.plot,ci.plot,scpi.plot,scm.ci.plot,
                    labels = c("A","B","C","D","E","F"), nrow = 3, ncol = 2)
ggsave("figures/BACIplot_Cleveland.png", plot = baci.plot.2)
ggsave("figures/Difference plot.png", plot = diffplot)
ggsave("figures/Weight plot.png", plot = weightplot, height = 6, width = 12)
ggsave("figures/Placebos plot.png", plot = placebos)
ggsave("figures/MPSE ratio plot.png", plot = mpseratio)
ggsave("figures/Method comparison plot.png", plot = trends.figure, height = 12, width = 11)
```

# Citations
```{r}
citation()
devtools::session_info()

c("tidyverse", "tidysynth","nlme","scpi") %>%
  map(citation) %>%
  print(style = "text")
```



