---
title: "GBIF Data Extraction"
author: "Asia Kaiser"
date: "2024-06-05"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(sf)
library(tidyverse)
library(iNEXT)
library(data.table)
library(arrow)
library(units)
```

# Load in Data
```{r}
rm(list = ls())
setwd("/Users/aska1693/Downloads/Chapter2_SyntheticControl/SyntheticControl-Repo/data_cleaning_code")
#test dataset
cities.test <- read.csv("raw_data/cities.test.csv")

#GBIF Hymenoptera data
gbif.all <- read.csv("raw_data/Gbif_All.csv", header=T, na.strings=c("","NA"))

#GBIF All taxa data
setDTthreads(parallel::detectCores())
gbif.all.taxa <- fread("raw_data/0001517-251009101135966_GBIF_ALL.csv", 
            header=T, na.strings=c("","NA"),
            showProgress = TRUE,
            select = c("decimalLatitude","decimalLongitude","recordedBy","year")) #fast read
head(gbif.all.taxa)
#write_parquet(gbif.all.taxa, "GBIF_all_taxa.parquet") #converting to Parquet for future use
#gbif.all.taxa <- read.csv("raw_data/0001517-251009101135966_GBIF_ALL.csv", header=T, na.strings=c("","NA"))
```

# Data Cleaning

## Removing non-target taxa
* Filtering data to just keep bees
* Inputting *genus spp* in all blank cells in the species column
```{r}
gbif <- gbif.all %>% 
  filter(gbif.all$family== c("Halictidae","Apidae","Colletidae","Melittidae","Megachilidae","Adrenidae"))%>%
  mutate(species = ifelse(is.na(species), paste0(genus, " spp"), species))
```

Code for creating smaller testing dataset
#```{r}
gbif.sub <- gbif[sample(nrow(gbif), 10000), ]
#```

## Assigning observations to cities

* Convert gbif data to an sf object
* Create buffer distances
```{r}
#Convert gbif to sf points
sf_data <- st_as_sf(gbif, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

#Convert cities.test to sf points
cities.sf <- st_as_sf(cities.test, coords = c("Long", "Lat"), crs = 4326)

#Create 45 km buffers around all cities
buffer45_distance <- set_units(45000, "m")  # 45 km in meters
cities.buffer <- st_buffer(cities.sf, dist = buffer45_distance)
```


Assign all observations to a city
```{r}
# Perform spatial join: assign City to points within buffer
sf_data <- st_join(sf_data, cities.buffer["City"], join = st_within)

# Fill points outside any city buffer
sf_data$City45km <- sf_data$City            # Copy joined column
sf_data$City45km[is.na(sf_data$City45km)] <- "Outside City"  # Default value
sf_data$City <- NULL  # Optional cleanup

# Convert back to data frame if needed
gbif <- as.data.frame(sf_data)

# Check unique city assignments
unique(gbif$City45km)
```

##Group Summaries

### Abundance values
```{r}
## Group Summaries

# Abundance for all bees (all families)
gbif.obs.all <- gbif %>%
  group_by(City45km, year) %>%
  summarise(n.obs45km = n(), .groups = "drop")

# Merge into cities.test
cities.test <- cities.test %>%
  pivot_longer(
    cols = starts_with("X"),
    names_to = "year",
    names_prefix = "X",
    values_to = "Population"
  ) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(gbif.obs.all, by = c("City" = "City45km", "year")) %>%
  mutate(n.obs45km = replace_na(n.obs45km, 0)
  )

# Quick check
summary(cities.test)
```

## Getting No of Unique Observors for Dataset with all taxa

* Convert gbif data to an sf object
```{r}
#Convert gbif to sf points
sf_data <- st_as_sf(gbif.all.taxa, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
```

Assign all observations to a city
```{r}
# Set Chunk parameters
chunk_size <- 500000   # 0.5M rows per chunk to fit in RAM
n_chunks <- ceiling(nrow(gbif.all.taxa) / chunk_size)

# Process chunks
for (i in 1:n_chunks) {
  cat("Processing chunk", i, "of", n_chunks, "\n")
  
  # Extract current chunk
  start_row <- (i - 1) * chunk_size + 1
  end_row <- min(i * chunk_size, nrow(gbif.all.taxa))
  gbif_chunk <- gbif.all.taxa[start_row:end_row, ]
  
  # Convert chunk to sf
  sf_chunk <- st_as_sf(gbif_chunk, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
  
  # Vectorized spatial join
  sf_chunk <- st_join(sf_chunk, cities.buffer["City"], join = st_within)
  
  # Assign City45km column
  sf_chunk$City45km <- sf_chunk$City
  sf_chunk$City45km[is.na(sf_chunk$City45km)] <- "Outside City"
  sf_chunk$City <- NULL
  
  # Convert to data.frame and drop geometry before saving
  sf_df <- as.data.frame(sf_chunk)
  sf_df$geometry <- NULL
  
  # Save chunk to Parquet
  chunk_file <- paste0("gbif_chunk_", i, ".parquet")
  write_parquet(sf_df, chunk_file)
  
  # Free memory
  rm(sf_chunk, gbif_chunk, sf_df)
  gc()
}

# Combine all Parquet chunks
file_list <- list.files(pattern = "gbif_chunk_.*\\.parquet$")
gbif.all.taxa <- lapply(file_list, read_parquet) %>% bind_rows()

# Check results
unique(gbif.all.taxa$City45km)
```

*Group summaries*
```{r}
# Number of unique observers (all taxa)
gbif.obs.observers <- gbif.all.taxa %>%
  group_by(City45km, year) %>%
  summarise(unique_observers = n_distinct(recordedBy), .groups = "drop")

# Merge into cities.test
cities.test <- cities.test %>%
  left_join(gbif.obs.observers, by = c("City" = "City45km", "year")) %>%
  mutate(unique_observers = replace_na(unique_observers, 0)
  )
```

#```{r}
# Aggregating data by year and city
gbif.obs.45km <- gbif %>%
  group_by(City45km, year) %>%
  summarise(n.obs45km = n(), .groups = 'drop')

# Transforming the cities data and merging with aggregated data
cities.test <- cities.test %>%
  pivot_longer(
    cols = starts_with("X"),   # Select columns that start with "X"
    names_to = "year",         # Name of the new "year" column
    names_prefix = "X",        # Remove the "X" prefix from the column names
    values_to = "Population"   # Name of the new "population" column
  ) %>%
  mutate(year = as.numeric(year)) %>%
  left_join(gbif.obs.45km, by = c("City" = "City45km", "year")) %>%
  mutate(n.obs45km = replace_na(n.obs45km, 0))

# View the transformed and merged data frame
summary(cities.test)
#```

### Extracting Biodiversity Values

* Summarize species counts per site
* Pivot dataframe wide to get into correct format for iNEXT
* Run iNEXT for estimated richness and diversity values
* Add values back to city dataframe

#```{r}

# Prepare gbif data and convert to wide format
gbif.wide.45 <- gbif %>%
  filter(City45km != "Outside City" & year >= min(cities.test$year)) %>%
  unite("year_city45km", year, City45km, sep = "_") %>%
  count(year_city45km, species) %>%
  pivot_wider(names_from = year_city45km, values_from = n, values_fill = 0)

# Prepare iNEXT input data
gbif.iNEXTinput.45 <- as.data.frame(gbif.wide.45[,-1])

# Run iNEXT to find richness data
gbif.iNEXT.45 <- iNEXT(gbif.iNEXTinput.45, q = 0, datatype = "abundance")

# Extract richness and diversity data
gbif.iNEXTvals.45km <- gbif.iNEXT.45$AsyEst %>%
  filter(Diversity %in% c("Species richness", "Shannon diversity")) %>%
  select(Assemblage, Diversity, Estimator) %>%
  pivot_wider(names_from = Diversity, values_from = Estimator, names_prefix = "est_") %>%
  separate(Assemblage, c("year", "City"), sep = "_") %>%
  rename(rich45km = `est_Species richness`, shann45km = `est_Shannon diversity`) %>%
  mutate(year = as.numeric(year))

# Merge with the original city dataset
cities.test <- cities.test %>%
  left_join(gbif.iNEXTvals.45km, by = c("City", "year"))%>%
  mutate(rich45km = replace_na(rich45km, 0))%>%
  mutate(shann45km = replace_na(shann45km, 0))

# View the updated cities.test data frame
head(cities.test)
#```

#Further narrowing donor pool
* Filter out cities experiencing treatment or facing abnormal shock
* Subsetting to cities with close number of observations
```{r}
#Subset for cities with closer number of observations during treatment year
cities.pool <- cities.test %>%
  filter(year == 2021) %>%
  filter(n.obs45km > 75 & n.obs45km < 500)

#Extract the city column
cities.pool <-cities.pool$City

#Keeping subset of cities in final dataframe
#Removing Boston, Atlanta and Baltimore for experiencing this treatment. Removing Houston for abnormal abundance spike in 2016.
cities.test <- cities.test %>%
  filter(City != "Boston"& City != "Atlanta" & City != "Baltimore" & City != "Houston")%>%
  filter(City %in% cities.pool)
```

Adding Socioeconomic variables
```{r}
setwd("/Users/aska1693/Downloads/Chapter2_SyntheticControl/SyntheticControl-Repo/data_cleaning_code")
socio.var <- read.csv("raw_data/SCM_socioeconomic - predictors.csv")

summary(socio.var)
socio.var <- socio.var %>%
    select(City,Median_HH_Income,PovertyIndex,BA_Population,MedianHomeValue)

cities.test <- left_join(cities.test, socio.var, by = "City")
```

# Save dataframe with summarized GBIF variables
```{r}
#Save processed data files in output folder
setwd("/Users/aska1693/Downloads/Chapter2_SyntheticControl/SyntheticControl-Repo")
write.csv(cities.test, "data/cities.scm.input.csv", row.names = FALSE)
```
